### 2021-10-29 manuscript/hiv-gwas.tex

\documentclass[]{article}
\usepackage[margin=1in]{geometry}
\usepackage{biblatex}
\addbibresource{references.bib}
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage{longtable} % for table that spans multiple pages

@@

	% Variance 
	\begin{scope}[xshift = 0cm, yshift = 0cm]
	\node[anchor = west] (alabel) at (-1, 4) {a) Variance partitioning};
	\node[constnode] (H2h) at (0, 0) {0.25};
	\node[constnode] (varz) at ($(H2h) + (1.8, 0)$) {$\sigma^2_z$};
	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {$H^2_{\bar{t}}$};
	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{g_h}$};
	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{\epsilon}$};
	\draw[taro] (H2h) -- (varh);
	\draw[taro] (varz) -- (varh);
	\draw[taro] (varz) -- (varv);
	\draw[taro] (H2) -- (varv);
	\draw[taro] (varh) -- (vare);
	\draw[taro] (varv) -- (vare);
	\draw[taro] (varz) -- (vare);
	\end{scope}

	% Virus
	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
	\node[anchor = west] (blabel) at (-1, 4.5) {b) Viral part of spVL};
	\node[constnode] (theta) at (0, 1.5) {$\theta$};
	\node[constnode] (g0) at (1.8, 1.5) {$g_{0}$};
	\node[constnode] (alpha) at (3.6, 1.5) {$\alpha$};
	\node[detnode] (sigma) at (5.4, 1.5) {$\sigma$};
	\node[constnode] (tree) at (7.2, 1.5) {$\Psi$};
	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
	\node[snode] (v) at (3.6, 3.25) {$\vec{g_v}$};
	\draw[taro] (theta) -- (v);
	\draw[taro] (g0) -- (v);
	\draw[taro] (alpha) -- (v);
	\draw[taro] (sigma) -- (v);
	\draw[taro] (tree) -- (v);
	\draw[taro] (varv) -- (sigma);
	\draw[taro] (tbar) -- (sigma);
	\draw[taro] (tbar) -- (tree);
	\draw[taro] (alpha) -- (sigma);
	\node[white] (varpic) at (3.6+0.9, -1.75) {};
	\draw[taro] (varpic) -- (varv);
	\node (vdist) at ($(v) + (1.8, 0.3)$) {Normal};
	\draw (v) -- (vdist);
	\draw[decoration={brace, mirror, raise=18pt}, sloped, decorate] (7.2, 1.5 - .5) -- node[below = 18pt] {OU parameters} (7.2, 1.5 + .5);
	\end{scope}

	% Host 
	\begin{scope}[xshift = 0cm, yshift = -9cm]
	\node[anchor = west] (clabel) at (-1, 5.55) {c) Host part of spVL};
	\node[constnode] (d) at (1.8, 1.5) {2};
	\node[detnode] (p) at (3.6, 1.5) {$p$};
	\node[snode] (G) at (1.8, 3) {$G$};
	\node[rectangle, dashed, very thick, inner sep=4mm, draw=black!100, fit = (G) ] (Gplate) {};
	\node[anchor=north west, inner sep=3pt] at (Gplate.north west) {\tiny $N\times M$};
	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{g_h}$};
	\node[constnode] (delta) at (3.6, 3) {$\delta$};
	\node[detnode] (h) at (3.6, 4.5) {$\vec{g_h}$};
	\node[constnode] (M) at (5.4, 3) {$M$};
	\draw[taro] (d) -- (G);
	\draw[taro] (p) -- (G);
	\draw[taro] (varh) -- (p);
	\draw[taro] (delta) -- (p);
	\draw[taro] (M) -- (p);
	\draw[taro] (M) -- (h);
	\draw[taro] (delta) -- (h);
	\draw[taro] (G) -- (h);
	\node[white] (varpic) at (3.6, -1.5) {};
	\draw[taro] (varpic) -- (varh);
	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Normal};
	\draw (h) -- (hdist);
	\node (gdist) at ($(G) + (-2, 0.6)$) {Binomial};
	\draw (G) -- (gdist);
	\end{scope}

	% Environment
	\begin{scope}[xshift = 10cm, yshift = -6.5cm]
	\node[anchor = west] (dlabel) at (-1, 3) {d) Environmental part of spVL};
	\node[detnode] (vare) at (0, 0) {$\sigma^2_{\epsilon}$};
	\node[constnode] (zero) at ($(vare) + (1.8, 0)$) {0};
	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{\epsilon}$};
	\draw[taro] (zero) -- (e);
	\draw[taro] (vare) -- (e);
	\node[white] (varpic) at (0, -1.5) {};
	\draw[taro] (varpic) -- (vare);
	\node (edist) at ($(e) + (1.8, 0.3)$) {Normal};
	\draw (e) -- (edist);
	\end{scope}

@@

\begin{table}[H]
	\begin{center}
		\begin{tabular}{cccc} \hline 
			Parameter & MLE & Maximum posterior probability & 95\% HPD \\ \hline 
			${g_{0}}$ & 4.17 & 4.23 & 2.73, 5.88 \\
			$\theta$ & 4.32 & 4.33 & 4.25, 4.42 \\
			$\sigma$ & 5.15 & 4.98 & 3.45, 6.61 \\
			$\alpha$ & 45.46 & 43.98 & 23.80, 63.36 \\
			$\sigma_e$ & 0.55 & 0.56 & 0.49, 0.63 \\ 
			$H^2_{\bar{t}}$ & 0.49 & 0.47 & 0.33, 0.60 \\ \hline
		\end{tabular}
		\caption{Inferred POUMM parameters.}
		\label{tab:POUMM-parameters}
	\end{center}
\end{table}

@@

\begin{table}[H]
    \begin{center}
    \begin{tabular}{lc}
    \hline
    Sample filter & Number of samples remaining \\
    \hline
    Subtype B pol sequences & 1516 \\
    With paired spVL measurement & 1516 \\
    $>$ 750 characters in sequence & 1493 \\
    Individual is of European ancestry & 1396 \\
    Kinship coefficient $>$ 0.09375 & TODO \\ 
    \hline
    \end{tabular}
    \caption{Number of samples for GWAS after sequential filtering steps.}
    \label{tab:sample-filtering}
    \end{center}
\end{table}

\begin{table}[H]
	\begin{center}
		\begin{tabular}{lc}
			\hline
			Variant filter & Number of variants remaining \\
			\hline 
			Raw data & 76979521 \\
			Missing genotype rate $>$ 0.01 & 7452957 \\
			Hardy-Weinburg exact test p-value $<$ 5x10$^-5$ & 7452425 \\
			Minor allele frequency $<$ 0.01 & 4556770 \\ \hline			
		\end{tabular}
		\caption{Number of variants for GWAS after sequential filtering steps.}
		\label{tab:variant-filtering}
	\end{center}
\end{table}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\linewidth]{figures_archived/200207_qq_plots.png}
	\caption{Quartile-quartile plots from association tests.}
	\label{fig:qq-plots}
\end{figure}

@@

\begin{longtable}[H]{p{3cm}p{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}

@@

\begin{table}[H]
	\begin{center}
		\begin{tabular}{cccl}
			\toprule 
			Parameter & Value & Uncertainty & Reference \\
			\midrule 
			$v_0$ & 5.54 & (4.04 - 7.25) HPD & \citealt{Mitov2018} \\
			$\theta$ & 4.45 & (4.41 - 4.49) HPD& \citealt{Mitov2018}  \\
			$\theta$ &  4.0 & (1.6 - 4.3) 95\% CI & \citealt{Bertels2018} \\
			$\theta$ & 4.1 & (3.5 - 4.9) 95\% CI & \citealt{Blanquart2017} \\
			$\alpha$ & 28.78 & (16.64 - 46.93) HPD & \citealt{Mitov2018} \\
			$\alpha$ & 32.7 & (0.03 - 57.6) 95\% CI & \citealt{Bertels2018} \\
			$\alpha$ & 7.6 & (1.2 - 10) 95\% CI & \citealt{Blanquart2017}** \\
			$\sigma$ & 2.97 & (1.95 - 4.37) HPD & \citealt{Mitov2018} \\
			$\sigma$ & 1.3 & (0.66 -1.87) 95\% CI &  \citealt{Blanquart2017} \\
			$\sigma_e$ & 0.77 & (0.73, 0.8) HPD & \citealt{Mitov2018} \\
			$\sigma_e$ & 0.61 & (0.54, 0.65) 95\% CI &  \citealt{Blanquart2017} \\
			\bottomrule 
		\end{tabular}
		\caption{POUMM parameters previously inferred from HIV data. Notes for each study: \protect\citet{Mitov2018} = 8,483 UK HIV cohort individuals, pol tree; \protect\citet{Bertels2018} = 3,036 SHCS individuals, pol tree; \protect\citet{Blanquart2017} = 1,581 subtype B individuals from BE, CH, FR, NL, UK, whole genome tree, **limited $\alpha$ to $\le$ 10}
		\label{tab:POUMMparams}
	\end{center}
\end{table}

\begin{table}[H]
	\begin{center}
			\begin{tabular}{ccl} \toprule 
			Measurement & Value & Reference \\ \midrule 
			$\bar{z}$ & $\approx$ 4.5 & \citealt{Mitov2018} \\
			$\bar{z}$ &  4.4 & \citealt{Blanquart2017} \\
			$\bar{z}$ & $\approx$ 4.5 & \citealt{Bonhoeffer2015} \\
			$\sigma^2_z$ &  0.73 &  \citealt{Mitov2018} \\
			$\sigma^2_z$ &  0.50 & \citealt{Blanquart2017} \\
			$\sigma^2_z$ & $\approx$ 0.5 & \citealt{Bonhoeffer2015} \\ \bottomrule 
		\end{tabular}
		\caption{Summary statistics for log spVL in previously sampled populations. $\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$). Values from \protect\citet{Blanquart2017, Mitov2018} are empirical while values from \protect\cite{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
		\label{tab:spVLmeasurements}
	\end{center}
\end{table}
%\todo{I guessed at bar(z) from figure in Mitov2018, search supplement for exact}

\begin{figure}[H]
	\includegraphics[width=\linewidth]{images/HIV_tree_branch_lengths}
	\caption{Left plot: branch length distributions for UK HIV tree and our simulated HIV tree prior to pruning down to 500 tips. The mean root-to-tip distance for the UK tree is 0.14 and the mean branch length is 0.02. The mean root-to-tip distance for the full simulated HIV tree is 0.14 and the mean branch length is 0.01. Right plot: branch length distribution for the tree of 500 tips used for simulation. The mean root-to-tip distance for the pruned HIV tree is 0.14 and the mean branch length is 0.03.}
	\label{fig:branchlengths}
\end{figure}

\end{doublespace}

>>>>

\documentclass[]{article}
\usepackage[margin=1in]{geometry}
\usepackage{biblatex}
\addbibresource{references.bib}
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage{longtable} % for table that spans multiple pages
\usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions

@@

@@

	% Variance 
	\begin{scope}[xshift = 0cm, yshift = 0cm]
	\node[anchor = west] (alabel) at (-1, 4) {A. Variance partitioning};
	\node[constnode] (H2h) at (0, 0) {0.25};
	\node[constnode] (varz) at ($(H2h) + (1.8, 0)$) {$\sigma^2_z$};
	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {$H^2_{\bar{t}}$};
	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{g_h}$};
	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{\epsilon}$};
	\draw[taro] (H2h) -- (varh);
	\draw[taro] (varz) -- (varh);
	\draw[taro] (varz) -- (varv);
	\draw[taro] (H2) -- (varv);
	\draw[taro] (varh) -- (vare);
	\draw[taro] (varv) -- (vare);
	\draw[taro] (varz) -- (vare);
	\end{scope}

	% Virus
	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
	\node[anchor = west] (blabel) at (-1, 4.5) {B. Viral part of spVL};
	\node[constnode] (theta) at (0, 1.5) {$\theta$};
	\node[constnode] (g0) at (1.8, 1.5) {$g_{0}$};
	\node[constnode] (alpha) at (3.6, 1.5) {$\alpha$};
	\node[detnode] (sigma) at (5.4, 1.5) {$\sigma$};
	\node[constnode] (tree) at (7.2, 1.5) {$\Psi$};
	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
	\node[snode] (v) at (3.6, 3.25) {$\vec{g_v}$};
	\draw[taro] (theta) -- (v);
	\draw[taro] (g0) -- (v);
	\draw[taro] (alpha) -- (v);
	\draw[taro] (sigma) -- (v);
	\draw[taro] (tree) -- (v);
	\draw[taro] (varv) -- (sigma);
	\draw[taro] (tbar) -- (sigma);
	\draw[taro] (tbar) -- (tree);
	\draw[taro] (alpha) -- (sigma);
	\node[white] (varpic) at (3.6+0.9, -1.75) {};
	\draw[taro] (varpic) -- (varv);
	\node (vdist) at ($(v) + (1.8, 0.3)$) {Normal};
	\draw (v) -- (vdist);
	\draw[decoration={brace, mirror, raise=18pt}, sloped, decorate] (7.2, 1.5 - .5) -- node[below = 18pt] {OU parameters} (7.2, 1.5 + .5);
	\end{scope}

	% Host 
	\begin{scope}[xshift = 0cm, yshift = -9cm]
	\node[anchor = west] (clabel) at (-1, 5.55) {C. Host part of spVL};
	\node[constnode] (d) at (1.8, 1.5) {2};
	\node[detnode] (p) at (3.6, 1.5) {$p$};
	\node[snode] (G) at (1.8, 3) {$G$};
	\node[rectangle, dashed, very thick, inner sep=4mm, draw=black!100, fit = (G) ] (Gplate) {};
	\node[anchor=north west, inner sep=3pt] at (Gplate.north west) {\tiny $N\times M$};
	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{g_h}$};
	\node[constnode] (delta) at (3.6, 3) {$\delta$};
	\node[detnode] (h) at (3.6, 4.5) {$\vec{g_h}$};
	\node[constnode] (M) at (5.4, 3) {$M$};
	\draw[taro] (d) -- (G);
	\draw[taro] (p) -- (G);
	\draw[taro] (varh) -- (p);
	\draw[taro] (delta) -- (p);
	\draw[taro] (M) -- (p);
	\draw[taro] (M) -- (h);
	\draw[taro] (delta) -- (h);
	\draw[taro] (G) -- (h);
	\node[white] (varpic) at (3.6, -1.5) {};
	\draw[taro] (varpic) -- (varh);
	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Normal};
	\draw (h) -- (hdist);
	\node (gdist) at ($(G) + (-2, 0.6)$) {Binomial};
	\draw (G) -- (gdist);
	\end{scope}

	% Environment
	\begin{scope}[xshift = 10cm, yshift = -6.5cm]
	\node[anchor = west] (dlabel) at (-1, 3) {D. Environmental part of spVL};
	\node[detnode] (vare) at (0, 0) {$\sigma^2_{\epsilon}$};
	\node[constnode] (zero) at ($(vare) + (1.8, 0)$) {0};
	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{\epsilon}$};
	\draw[taro] (zero) -- (e);
	\draw[taro] (vare) -- (e);
	\node[white] (varpic) at (0, -1.5) {};
	\draw[taro] (varpic) -- (vare);
	\node (edist) at ($(e) + (1.8, 0.3)$) {Normal};
	\draw (e) -- (edist);
	\end{scope}

@@

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{figures_archived/200207_qq_plots.png}
	\caption{Quartile-quartile plots from association tests.}
	\label{fig:qq-plots}
\end{figure}

\end{doublespace}  % don't doublspace tables

@@

\begin{table}[H]
\caption{POUMM parameter estimates for spVL based on Swiss data. HPD = Highest posterior density.}
	\begin{tabular}{lccc} \hline 
	Parameter & Posterior mean & 95\% HPD \\ \hline 
	${g_{0}}$ & 4.23 & (1.72, 6.71) \\
	$\theta$ & 4.47 & (4.37, 4.58) \\
	$\sigma$ & 5.25 & (2.37, 7.9) \\
	$\alpha$ & 57.65 & (19.49, 95.2) \\
	$\sigma_e$ & 0.54 & (0.43, 0.65) \\ 
	$H^2_{\bar{t}}$ & 0.45 & (0.24, 0.67) \\ \hline
	\end{tabular}
	\label{tab:POUMM-parameters}
\end{table}

\begin{table}[H]
\caption{POUMM parameter estimates for spVL from previous studies.}
	\begin{tabular}{lcccp{0.35\linewidth}}
	\hline 
	Parameter & Value & Uncertainty & Reference & Notes \\
	\hline 
	$v_0$ & 5.54 & (4.04 - 7.25) HPD & \citealt{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
	$\theta$ & 4.45 & (4.41 - 4.49) HPD & \citealt{Mitov2018} & \\
	$\theta$ &  4.0 & (1.6 - 4.3) 95\% CI & \citealt{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
	$\theta$ & 4.1 & (3.5 - 4.9) 95\% CI & \citealt{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
	$\alpha$ & 28.78 & (16.64 - 46.93) HPD & \citealt{Mitov2018} & \\
	$\alpha$ & 32.7 & (0.03 - 57.6) 95\% CI & \citealt{Bertels2018} & \\
	$\alpha$ & 7.6 & (1.2 - 10) 95\% CI & \citealt{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
	$\sigma$ & 2.97 & (1.95 - 4.37) HPD & \citealt{Mitov2018} & \\
	$\sigma$ & 1.3 & (0.66 -1.87) 95\% CI &  \citealt{Blanquart2017} & \\
	$\sigma_e$ & 0.77 & (0.73, 0.8) HPD & \citealt{Mitov2018} & \\
	$\sigma_e$ & 0.61 & (0.54, 0.65) 95\% CI &  \citealt{Blanquart2017} & \\
	\hline 
	\end{tabular}
	\label{tab:POUMMparams}
\end{table}

\begin{table}[H]
\caption{Summary statistics for log spVL in previously sampled populations. $\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$). Values from \protect\citet{Blanquart2017, Mitov2018} are empirical; values from \protect\cite{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
	\begin{tabular}{lcc} \hline 
	Measurement & Value & Reference \\ \hline 
	$\bar{z}$ & $\approx$ 4.5 & \citealt{Mitov2018} \\
	$\bar{z}$ &  4.4 & \citealt{Blanquart2017} \\
	$\bar{z}$ & $\approx$ 4.5 & \citealt{Bonhoeffer2015} \\
	$\sigma^2_z$ &  0.73 &  \citealt{Mitov2018} \\
	$\sigma^2_z$ &  0.50 & \citealt{Blanquart2017} \\
	$\sigma^2_z$ & $\approx$ 0.5 & \citealt{Bonhoeffer2015} \\ \hline 
	\end{tabular}
	\label{tab:spVLmeasurements}
\end{table}
%\todo{I guessed at bar(z) from figure in Mitov2018, search supplement for exact}

\begin{table}[H]
\caption{Number of samples for GWAS after sequential filtering steps.}
    \begin{tabular}{lc}
    \hline
    Sample filter & Number of samples remaining \\
    \hline
    Subtype B pol sequences & 1516 \\
    With paired spVL measurement & 1516 \\
    $>$ 750 characters in sequence & 1493 \\
    Individual is of European ancestry & 1396 \\
    Kinship coefficient $>$ 0.09375 & 1392 \\ 
    \hline
    \end{tabular}
    \label{tab:sample-filtering}
\end{table}

\begin{table}[H]
\caption{Number of variants for GWAS after sequential filtering steps.}
	\begin{tabular}{lc}
	\hline
	Variant filter & Number of variants remaining \\
	\hline 
	Raw data & 76979521 \\
	Missing genotype rate $>$ 0.01 & 7452957 \\
	Hardy-Weinburg exact test p-value $<$ 5x10$^-5$ & 7452425 \\
	Minor allele frequency $<$ 0.01 & 4556770 \\ \hline			
	\end{tabular}
	\label{tab:variant-filtering}
\end{table}

\begin{longtable}[H]{lp{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}

@@

% \begin{figure}[H]
% 	\includegraphics[width=\linewidth]{images/HIV_tree_branch_lengths}
% 	\caption{Left plot: branch length distributions for UK HIV tree and our simulated HIV tree prior to pruning down to 500 tips. The mean root-to-tip distance for the UK tree is 0.14 and the mean branch length is 0.02. The mean root-to-tip distance for the full simulated HIV tree is 0.14 and the mean branch length is 0.01. Right plot: branch length distribution for the tree of 500 tips used for simulation. The mean root-to-tip distance for the pruned HIV tree is 0.14 and the mean branch length is 0.03.}
% 	\label{fig:branchlengths}
% \end{figure}
